---
title: "demo_ChromTRAP"
author: "Kaho Itoh"
date: "2026-06-25"
output: 
  github_document:
      toc: true
---

# Demo ChromTRAP analysis

This demo reproduces the key ChromTRAP-based differential expression analysis
using a downsampled Seurat object.

The demo object, `SO_demo.rds`, contains cells from the BLA region only.  
Specifically, it includes excitatory neuron subtypes from the basal amygdala
(BA) and lateral amygdala (LA). This reduced object is provided so that the
main analysis can be run quickly without requiring the full Seurat object.

In this example, cells are classified by AP-1 / FOS motif activity using the
chromVAR deviation score for motif `MA1141.2`. Cells in the top quantile are
defined as motif-high cells, and cells in the bottom quantile are defined as
motif-low cells within each sample group.

# Load libraries

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(dplyr)
```

# Config

```{r config}
percentile_high <- 0.9
percentile_low <- 0.1
```

# Load demo data

```{r load-data}
SO <- readRDS("SO_demo.rds")
```

The demo object contains the following number of cells per annotation and sample:

```{r check-demo-object}
table(SO@meta.data$Annotation, SO@meta.data$Sample)
```

The AP-1 / FOS motif activity score is extracted from the chromVAR assay using
motif ID `MA1141.2`.

```{r extract-chromvar-score, warning=FALSE, message=FALSE}
chromvar_mat <- GetAssayData(
  object = SO,
  assay = "chromvar",
  slot = "data"
)

chromvar_count_selected <- chromvar_mat["MA1141.2", ]
```

# Define motif-high and motif-low cells

Cells are ranked by chromVAR score within each sample group.  
Here, the top 10% of cells are defined as motif-high cells, and the bottom 10%
are defined as motif-low cells.

```{r define-high-low-cells}
df_chromvar_count <- data.frame(
  ChromVAR_count = as.numeric(chromvar_count_selected),
  CellID = names(chromvar_count_selected),
  HC_Cond4h = SO@meta.data[names(chromvar_count_selected), "Sample"]
)

df_High <- df_chromvar_count %>%
  group_by(HC_Cond4h) %>%
  mutate(q = quantile(ChromVAR_count, percentile_high, na.rm = TRUE)) %>%
  dplyr::filter(ChromVAR_count >= q)

df_Low <- df_chromvar_count %>%
  group_by(HC_Cond4h) %>%
  mutate(q = quantile(ChromVAR_count, percentile_low, na.rm = TRUE)) %>%
  dplyr::filter(ChromVAR_count <= q)
```


# Select cells for ChromTRAP comparison

In this demo, we compare Cond4h motif-high cells against HC motif-low cells.

```{r select-trapped-cells}
Cond_High_CellID <- df_High$CellID[df_High$HC_Cond4h == "Cond4h"]
HC_Low_CellID <- df_Low$CellID[df_Low$HC_Cond4h == "HC"]

cells_use <- c(Cond_High_CellID, HC_Low_CellID)

SO_trapped <- subset(SO, cells = cells_use)

SO_trapped$ChromTRAP_group <- case_when(
  colnames(SO_trapped) %in% Cond_High_CellID ~ "Cond4h_High",
  colnames(SO_trapped) %in% HC_Low_CellID ~ "HC_Low"
)

table(SO_trapped$ChromTRAP_group)
```

# Differential expression analysis

Differential expression is performed between Cond4h motif-high cells and HC
motif-low cells using the SCT assay.

```{r find-markers, warning=FALSE, message=FALSE}
FM_res <- FindMarkers(
  object = SO_trapped,
  ident.1 = Cond_High_CellID,
  ident.2 = HC_Low_CellID,
  test.use = "wilcox",
  min.pct = 0.01,
  assay = "SCT"
)

head(FM_res)
```

